---
title: "Beyond peak intensity: classifying emotions from the dynamics of facial Action Units"
subtitle: "Machine learning analyses (Random Forest & C5.0)"
author: "XX, XX, XX, ..."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
  word_document:
    toc: true
    number_sections: true
bibliography: references.bib
link-citations: yes
csl: apa.csl
---

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5, fig.align = "center"
)
set.seed(123)
options(scipen = 999)
```

# About this document {-}

**Question.** Can the emotion elicited by a clip be recognised from the
*dynamics* of the facial Action Units (AUs) — not only how high AU peaks are,
but also how long they last and how fast they rise?

**Data.** One observation = one subjective intensity rating given by a
participant during a clip. For each of the 17 AUs, the predictors describe the
AU peak that was closest in time to the rating (see `01_preprocessing.R`,
steps 12–13):

| Feature suffix | Meaning |
|---|---|
| `_height` | peak height (maximum intensity, baseline-corrected) |
| `_duration` | peak duration in frames (end − start) |
| `_rise_slope` | height / (top − start): mean speed of the rising phase |

The outcome is the emotion category of the clip (10 classes).

**Procedure.**

1. 75 % / 25 % stratified split into a training set and an independent
   hold-out test set.
2. Random Forest [ranger, @ranger] and C5.0 decision trees / rules
   [@C50] trained with the caret package [@caret]; hyperparameters tuned with
   5-fold cross-validation repeated twice (5 × 2 CV) on the training set.
3. Evaluation on the hold-out test set (accuracy, confusion matrices).
4. Interpretation: variable importance (both models) and decision rules (C5.0).

**Important:** the order of the chunks must not be changed. Several steps use
random numbers (data split, forests, cross-validation folds): the results of
the article are reproduced only when the chunks are run in this order.

## Data analyses (text for the Methods section) {-}

Data analyses were conducted using R version 4.2.3 [@Rcore] within the RStudio
environment [@Rstudio]. All analyses were implemented using literate
programming [i.e., @rmarkdown]. Machine learning procedures were carried out
with the caret package [@caret], relying on the ranger algorithm for Random
Forest classification [@ranger] and the C5.0 algorithm for decision tree and
rule-based modelling [@C50]. Model training and evaluation were performed with
a 5 × 2 repeated cross-validation procedure and an independent 25 % hold-out
test set. Graphical visualisations were produced using the ggplot2 package
[@ggplot2].


# Setup

```{r packages}
source(here::here("R", "00_setup.R"))

library(caret)       # unified interface for training / tuning models
library(ranger)      # fast Random Forest
library(C50)         # C5.0 decision trees and rules
library(doParallel)  # parallel cross-validation
library(foreach)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(MLmetrics)   # multi-class metrics used by caret::multiClassSummary

save_figure <- function(plot, file, width = 8, height = 6, dpi = 600) {
  ggsave(file.path(paths$figures, file), plot = plot,
         width = width, height = height, dpi = dpi)
}
save_table <- function(df, file) {
  write.csv(df, file.path(paths$tables, file), row.names = FALSE)
}
```

# Data

## Import

```{r import}
data_path <- file.path(paths$processed, "ml_dataset.csv")
if (!file.exists(data_path)) {
  stop("File not found: ", data_path, "\nRun 01_preprocessing.R first.")
}

ml_raw <- read.csv(data_path, sep = ";", dec = ",", stringsAsFactors = FALSE) %>%
  harmonise_names()   # accepts the original file with French column names

str(ml_raw, list.len = 10)
table(ml_raw$condition)
```

## Selection of the predictors

Predictors = height, duration and rise slope of each AU's closest peak. The
time position of the peak (`_position`) and the identifiers are **not** used.

```{r predictors}
predictors <- grep("_(height|duration|rise_slope)$", names(ml_raw), value = TRUE)
length(predictors)

# Consistency check with the original analysis, in which predictors were
# selected by column position (columns 7:23 and 41:69 of the original file,
# recognisable by its first unnamed column "X").
if ("X" %in% names(ml_raw)) {
  legacy_idx <- c(7:23, 41:69)
  legacy <- names(ml_raw)[legacy_idx[legacy_idx <= ncol(ml_raw)]]
  if (!setequal(predictors, legacy)) {
    warning("Name-based and original index-based predictor sets differ:\n",
            "  only in name-based : ", paste(setdiff(predictors, legacy), collapse = ", "), "\n",
            "  only in index-based: ", paste(setdiff(legacy, predictors), collapse = ", "))
  }
}
```

## Preparation

```{r preparation}
df <- ml_raw[, c("ID", "Clip", "condition", predictors)]
df <- na.omit(df)

df$condition <- translate_emotions(df$condition)   # French -> English, ordered
df$ID   <- as.factor(df$ID)
df$Clip <- as.factor(df$Clip)

dim(df)
length(unique(df$ID))     # participants
length(unique(df$Clip))   # clips

p_classes <- ggplot(df, aes(x = condition)) +
  geom_bar() +
  theme_minimal() +
  labs(x = "Emotion category", y = "Number of observations",
       title = "Class distribution") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_classes
save_figure(p_classes, "Distribution_Class.png")
```

## Train / test split (75 % / 25 %)

`createDataPartition()` samples within each emotion, so that both sets have
the same class proportions (stratified split).

```{r split}
set.seed(123)
train_indices <- caret::createDataPartition(df$condition, p = 0.75, list = FALSE)
train_data <- df[train_indices, ]
test_data  <- df[-train_indices, ]

# Identifiers are removed from the predictors (they are not facial features)
train_x <- train_data %>% select(-ID, -Clip)
test_x  <- test_data  %>% select(-ID, -Clip)

nrow(train_x)
nrow(test_x)
round(prop.table(table(train_x$condition)), 3)
round(prop.table(table(test_x$condition)), 3)
```

> **Note on independence.** The split is made at the level of observations:
> the same participant (and the same clip) can appear in both the training and
> the test sets.


# Random Forest

## Initial model (default settings)

A first forest of 1000 trees, with 2 candidate predictors per split
(`mtry = 2`).

```{r rf-initial}
model_ranger <- ranger(
  condition ~ ., data = train_x,
  num.trees = 1000, mtry = 2,
  importance = "impurity", probability = TRUE
)

# Predicted class = the class with the highest predicted probability
pred <- predict(model_ranger, data = test_x)
pred_classes <- factor(
  colnames(pred$predictions)[apply(pred$predictions, 1, which.max)],
  levels = levels(train_x$condition)
)

cm_rf_initial <- caret::confusionMatrix(pred_classes, test_x$condition)
cm_rf_initial

p_cm <- ggplot(as.data.frame(cm_rf_initial$table),
               aes(x = Prediction, y = Reference, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme_minimal() +
  labs(title = "Confusion matrix \u2013 Random Forest",
       x = "Predicted emotion", y = "True emotion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_cm
save_figure(p_cm, "Figure_confusion_matrix.png")
```

## Variable importance

Importance = mean decrease in Gini impurity: how much, on average, splits on
a variable make the tree nodes "purer" (more homogeneous in emotion).

```{r rf-importance, fig.width=7, fig.height=6}
imp <- sort(model_ranger$variable.importance, decreasing = TRUE)
imp_df <- data.frame(Variable = names(imp), Importance = as.numeric(imp))

# Readable labels, e.g. "AU12 (Lip corner puller) – height"
feature_labels <- c(height = "height", duration = "duration", rise_slope = "rise slope")
imp_df <- imp_df %>%
  mutate(
    AU      = sub("^(AU[0-9]{2}).*$", "\\1", Variable),
    feature = feature_labels[sub("^AU[0-9]{2}_r_", "", Variable)],
    Label   = paste0(AU, " (", au_names[AU], ") \u2013 ", feature)
  ) %>%
  filter(is.finite(Importance)) %>%
  arrange(desc(Importance))

imp_top <- head(imp_df, 20)

p_imp <- ggplot(imp_top, aes(x = reorder(Label, Importance), y = Importance)) +
  geom_col() +
  coord_flip() +
  geom_text(aes(label = round(Importance, 2)), hjust = -0.1, size = 3) +
  labs(title = "Top variable importance scores (Random Forest)",
       x = NULL, y = "Variable importance (Gini impurity)") +
  theme_minimal(base_size = 11) +
  expand_limits(y = max(imp_top$Importance) * 1.1)
p_imp
save_figure(p_imp, "importance_scores.png")
```

## Tuning with 5 × 2 cross-validation

The training set is split into 5 folds; each fold is used once for
validation while the model is trained on the 4 others; the whole procedure is
repeated twice. We tune `min.node.size` (minimum number of observations in a
terminal node: small values = deeper trees).

```{r rf-cv-setup}
# Parallel computing (one core is left free)
cl <- parallel::makeCluster(max(1L, parallel::detectCores() - 1L))
doParallel::registerDoParallel(cl)

fitControl <- caret::trainControl(
  method = "repeatedcv", number = 5, repeats = 2,
  classProbs = TRUE,
  summaryFunction = caret::multiClassSummary,
  allowParallel = TRUE
)

tuneGrid <- expand.grid(
  mtry = 2,
  splitrule = "gini",
  min.node.size = c(1, 5, 10)
)
```

```{r rf-cv-train}
set.seed(123)
model_rf_cv <- caret::train(
  condition ~ ., data = train_x,
  method = "ranger",
  trControl = fitControl,
  tuneGrid = tuneGrid,
  num.trees = 500,
  importance = "none",
  respect.unordered.factors = "order"
)

model_rf_cv   # performance averaged over the 10 validation folds

ggplot(model_rf_cv$results, aes(x = factor(min.node.size), y = Accuracy, group = 1)) +
  geom_line() + geom_point(size = 2) +
  labs(x = "min.node.size", y = "CV accuracy (mean)",
       title = "Random Forest \u2014 cross-validation results") +
  theme_minimal(base_size = 11)
```

## Evaluation on the hold-out test set (25 %)

```{r rf-test}
pred_rf_cv <- predict(model_rf_cv, newdata = test_x)
cm_rf_cv <- caret::confusionMatrix(pred_rf_cv, test_x$condition)
cm_rf_cv

# Confusion matrix normalised by true class (each row sums to 100 %)
cm_norm <- as.data.frame(cm_rf_cv$table) %>%
  group_by(Reference) %>%
  mutate(Percent = Freq / sum(Freq)) %>%
  ungroup()

p_cm_norm <- ggplot(cm_norm, aes(x = Prediction, y = Reference, fill = Percent)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.1f%%", Percent * 100)), size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme_minimal(base_size = 12) +
  labs(title = "Normalized confusion matrix \u2013 Random Forest classifier",
       x = "Predicted emotion", y = "True emotion", fill = "Proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_cm_norm
save_figure(p_cm_norm, "Figure_RF_confusion_matrix_normalized.png")
```

## Final refit with the best hyperparameters (variable importance)

```{r rf-final}
best <- model_rf_cv$bestTune

rf_final <- ranger::ranger(
  condition ~ ., data = train_x,
  num.trees = 1000,
  mtry = best$mtry,
  splitrule = as.character(best$splitrule),
  min.node.size = best$min.node.size,
  importance = "impurity",
  respect.unordered.factors = "order",
  seed = 123
)
```


# C5.0

C5.0 builds decision trees (or sets of rules) that are directly readable,
which makes the classification easier to interpret than a Random Forest.

## Baseline model (caret default grid)

```{r c50-baseline}
model_c50_default <- caret::train(
  condition ~ ., data = train_x,
  method = "C5.0",
  trControl = fitControl       # same 5 × 2 CV as above
)

model_c50_default
model_c50_default$bestTune
caret::varImp(model_c50_default)
```

## Tuning with 5 × 2 cross-validation

Tuned hyperparameters:

- `model`: a single tree or a set of rules;
- `trials`: number of boosting iterations (1 = no boosting);
- `winnow`: whether uninformative predictors are discarded before fitting.

```{r c50-cv-setup}
fitControl_c50 <- caret::trainControl(
  method = "repeatedcv", number = 5, repeats = 2,
  classProbs = TRUE,
  summaryFunction = caret::defaultSummary,
  allowParallel = TRUE
)

tuneGrid_c50 <- expand.grid(
  model  = c("tree", "rules"),
  trials = c(1, 10, 30),
  winnow = c(TRUE, FALSE)
)
```

```{r c50-cv-train}
set.seed(123)
model_c50 <- caret::train(
  condition ~ ., data = train_x,
  method = "C5.0",
  trControl = fitControl_c50,
  tuneGrid = tuneGrid_c50,
  metric = "Accuracy"
)

model_c50

# Hold-out test evaluation (25 %)
pred_c50 <- predict(model_c50, newdata = test_x)
cm_c50 <- caret::confusionMatrix(pred_c50, test_x$condition)
cm_c50

# Parallel computing is no longer needed
parallel::stopCluster(cl)
foreach::registerDoSEQ()
```

## Variable importance

For C5.0, importance = percentage of training observations that fall in a
node where the variable is used to split.

```{r c50-importance}
vi_c50 <- caret::varImp(model_c50)
vi_c50
plot(vi_c50, top = 20)

vi_df <- vi_c50$importance
vi_df$Variable <- rownames(vi_df)
vi_df <- head(vi_df[order(vi_df$Overall, decreasing = TRUE), ], 20)

p_vi_c50 <- ggplot(vi_df, aes(x = reorder(Variable, Overall), y = Overall)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 20 Variable Importance (C5.0)", x = "Variables", y = "Importance") +
  theme_minimal()
p_vi_c50
save_figure(p_vi_c50, "variable_importance_c50.png", width = 10, height = 6, dpi = 300)
```

## Decision rules

The printed summary of the final C5.0 model lists its leaves/rules as
`condition: class (n/errors)`, where `n` is the number of training
observations covered and `errors` the number misclassified. We extract these
lines and keep the reliable ones: coverage ≥ 10 observations and error
rate ≤ 30 %.

```{r c50-rules}
model_summary <- capture.output(summary(model_c50$finalModel))

rules <- data.frame(rule = grep(": .*\\([0-9.]+", model_summary, value = TRUE)) %>%
  mutate(
    class      = str_extract(rule, paste(emotion_order, collapse = "|")),
    coverage   = as.numeric(str_match(rule, "\\(([0-9.]+)")[, 2]),
    errors     = as.numeric(str_match(rule, "\\([0-9.]+/([0-9.]+)\\)")[, 2]),
    errors     = ifelse(is.na(errors), 0, errors),
    error_rate = errors / coverage
  )

rules_selected <- rules %>%
  filter(coverage >= 10, error_rate <= .30) %>%
  arrange(class, error_rate, desc(coverage))
rules_selected
save_table(rules_selected, "C50_selected_rules.csv")

# Three best rules per emotion
top_rules <- rules_selected %>%
  group_by(class) %>%
  arrange(error_rate, desc(coverage)) %>%
  slice_head(n = 3) %>%
  ungroup()
save_table(top_rules, "C50_top_rules_by_emotion.csv")

# How often each AU appears in the selected rules of each emotion (top 5)
au_frequency <- rules_selected %>%
  mutate(AU = str_extract_all(rule, "AU[0-9]{2}")) %>%
  unnest(AU) %>%
  count(class, AU, sort = TRUE) %>%
  group_by(class) %>%
  slice_max(n, n = 5) %>%
  ungroup()
save_table(au_frequency, "C50_AU_frequency_by_emotion.csv")
```

```{r c50-rules-heatmap}
threshold <- 18   # above this count, labels are printed in white for contrast

p_rules <- ggplot(au_frequency, aes(x = AU, y = class, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n, color = n >= threshold), size = 3) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "white"), guide = "none") +
  scale_fill_gradientn(colors = c("#f7f7f7", "#6baed6", "#08306b")) +
  labs(title = "Frequency of Action Units Across C5.0 Decision Rules",
       x = "Action Units", y = "Emotion", fill = "Frequency") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold"))
p_rules
save_figure(p_rules, "heatmap_AU_emotions.png", width = 10, height = 6, dpi = 300)
```


# Model comparison

```{r comparison}
comparison <- data.frame(
  Model = c("Random Forest (initial)", "Random Forest (CV-tuned)", "C5.0 (CV-tuned)"),
  Test_accuracy = round(c(cm_rf_initial$overall["Accuracy"],
                          cm_rf_cv$overall["Accuracy"],
                          cm_c50$overall["Accuracy"]), 4)
)
comparison

# Five most important variables of each tuned model
top5_rf  <- names(sort(rf_final$variable.importance, decreasing = TRUE))[1:5]
top5_c50 <- rownames(vi_c50$importance)[order(-vi_c50$importance$Overall)][1:5]
```

## Summary of the results {-}

To identify which Action Units (AUs) best discriminate between emotion
categories, we trained two supervised classifiers: a Random Forest (ranger) and
a C5.0 decision tree/rule-based model. Both models were trained on 75 % of the
data and evaluated on an independent 25 % hold-out test set; hyperparameters
were optimised using 5 × 2 repeated cross-validation. On the hold-out set, the
Random Forest achieved an accuracy of
`r round(100 * cm_rf_cv$overall["Accuracy"], 2)` %, while the C5.0 model
reached `r round(100 * cm_c50$overall["Accuracy"], 2)` %. Within
cross-validation, the best C5.0 configuration obtained a mean accuracy of
`r round(100 * max(model_c50$results$Accuracy), 2)` % using the
`r model_c50$results$model[which.max(model_c50$results$Accuracy)]` setting.
Overall, both algorithms performed comparably, with the Random Forest offering
more stable estimates of variable importance, and the C5.0 model providing
interpretable decision rules. The five most discriminative features were
`r paste(top5_rf, collapse = ", ")` for the Random Forest model and
`r paste(top5_c50, collapse = ", ")` for the C5.0 model. The complete set of
rules and the confusion matrices are presented in Appendix 1.


# Appendix 1 {-}

```{r appendix, echo=FALSE}
cat("### Random Forest (ranger) \u2014 tuned refit\n")
print(rf_final)

cat("\n### C5.0 \u2014 cross-validated model\n")
print(summary(model_c50$finalModel))

cat("\n### Confusion matrix \u2014 Random Forest, CV-tuned (hold-out test)\n")
print(cm_rf_cv)

cat("\n### Confusion matrix \u2014 Random Forest, initial model (hold-out test)\n")
print(cm_rf_initial)

cat("\n### Confusion matrix \u2014 C5.0 (hold-out test)\n")
print(cm_c50)
```

# Session information {-}

```{r session}
sessionInfo()
```

# References {-}
